---
title: "Starbucks Food Menu Items Exploratory Data Analysis"
author: "Emi Strati"
date: "2023-06-04"
output:
  html_document:
    theme: united
    toc: true
    toc_float: true
    toc_collapsed: true
---

# Starbucks Nutrition Facts Exploratory Data Analysis

The "starbucks.csv" dataset contains nutrition facts for several Starbucks food items. The dataset used in this document was extracted from Kaggle (<https://www.kaggle.com/datasets/utkarshx27/starbucks-nutrition>).

# Preliminary

We will use the **tidyverse** package to explore the data and **ggplot2** to create the visualizations. The **tidyverse** package contains the **dplyr** package for data manipulation, among others.

```{r}
library(tidyverse)
library(ggplot2)

# importing the file
starbucks <- read.csv(file = 'starbucks.csv')
head(starbucks)
```

We can remove the X column, which serves as a secondary index.

```{r}
sbux <- subset(starbucks, select = -c(X))
str(sbux) # viewing the data structure
```

There are 77 observations in the dataset, described through the following 7 variables:

`item`: food item, string

`calories`: calories, numerical

`fat`: grams of fat in item, numerical

`carb`: grams of carbohydrates in item, numerical

`fiber`: grams of fiber in food item, numerical

`protein`: grams of protein in food item, numerical `type`: a factor with levels bakery, bistro box, hot breakfast, parfait, petite, salad, and sandwich

Finding out how many items there are of each type:

```{R}
count_type <- sbux %>% 
  group_by(type) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))

count_type

# pie chart of results
ggplot(count_type, aes(x = "", y = perc, fill = type)) +
  geom_col(color = NA) +
  geom_label(aes(label = labels),
             position = position_stack(vjust = 0.5),
             show.legend = FALSE) +
  guides(fill = guide_legend(title = "Food Types")) +
  coord_polar(theta = "y") + 
  theme_void() +
  labs(title = "Breakdown of Starbucks Food Menu Items by Type")
```

# Food Item Calories: Highest, Lowest & Outliers
One of the first things we can explore in this dataset are the foods with the highest and lowest amounts of calories. Since we have 77 items in our dataset, we can just focus on the top 5 foods with the most/least amounts of calories. We can do so using the `select()` function from dplyr.
```{r}
high_cal <- sbux %>% 
  arrange(desc(calories)) %>% 
  select(item, calories, type) %>% 
  head(5)
high_cal
```
The Sausage & Cheddar Classic Breakfast sandwich has the most calories out of all the items in the menu, but the next items come very close to it. Another interesting observation is that every item after the sandwich is of the bakery type. This might suggest that bakery items have a higher average caloric content than all other types. We will check this in the next section.
```{r}
low_cal <- sbux %>% 
  arrange(calories) %>% 
  select(item, calories, type) %>% 
  head(5) 
low_cal
```
The Deluxe Fruit Blend has noticeably less calories than the other low calorie items, suggesting that it may be an outlier in the dataset. Calorie outliers will be covered in a later section.

# Average Calorie Content of Each Food Item Type
In the previous section, we observed that 4 out of the 5 items with the most calories belong to the bakery food type. This raises the question: on average, how does each type of food compare, calorie wise?

To find out, we group the data by type, summarize it using the mean function, and arrange it in a descending order from highest average calories to lowest.
```{r}
avg_cal <- sbux %>% 
  group_by(type) %>% 
  summarise(avg_cal = mean(calories)) %>%
  arrange(desc(avg_cal))
avg_cal
```
It turns out that sandwiches are in fact the items containing the most calories, on average. We can visualize the data above using a bar chart:
```{r}
ggplot(avg_cal, aes(x=reorder(type, avg_cal), y=avg_cal, fill=type)) +
  geom_bar(stat='identity') +
  coord_flip() +
  labs(title='Average Calories by Type', x='Type', y='Average Calories')
```

# Outliers in calorie information
When exploring data, it is important to detect any possible outliers, since they can have a large influence on statistics derived from the dataset. We want to make sure that the average calorie data by type we extracted in the previous step is not skewed due to any extreme values. To find out about any outliers, we need to map the distribution of the data using a box plot.

```{r}
boxplot(starbucks$calories,
        names = c("Calories"),
        main = "Calorie Distribution in Starbucks Food Items")
```
There is only one outlier in the dataset, which lays on the lower extreme of the data. We can identify that particular food item using the code blow:
```{r}
c_out <- boxplot.stats(starbucks$calories)$out
c_out_index <- which(starbucks$calories %in% c_out)
c_out_item <- starbucks[c_out_index, "item"]
c_out_item
```
When we selected the bottom 5 items in terms of caloric content, Deluxe Fruit Blend was the item with the least calories. This confirmed our earlier assumption that this item was an outlier in the dataset. Deluxe Fruit Blend is also the only item that belongs to the “salad” type.
```{r}
low_cal <- sbux %>% 
  arrange(calories) %>% 
  select(item, calories, type) %>% 
  head(1) 
low_cal
```
# Food Items with the Highest Levels of Each Nutritional Property
Next, we can check which food items have the highest levels of each nutritional property: fat, carbohydrates, fiber, and protein, by using a similar expression to the one in the previous section.

a. Fat
```{r}
high_fat <- sbux %>% arrange(desc(fat)) %>% select(item, fat, type) %>% head(5)
high_fat
```
b. Carbohydrates
```{r}
high_carb <- sbux %>% arrange(desc(carb)) %>% select(item, carb, type) %>% head(5)
high_carb
```
c. Fiber
```{r}
high_fiber <- sbux %>% arrange(desc(fiber)) %>% select(item, fiber, type) %>% head(5)
high_fiber
```
d. Protein
```{r}
high_protein <- sbux %>% arrange(desc(protein)) %>% select(item, protein, type) %>% head(5)
high_protein
```

# Food item types with the highest average fat, carbohydrate, fiber, and protein content

In the previous section, we made two interesting observations:

1. Bakery items seem to have in general more carbohydrates than any other item.

2. Sandwiches seem to have the highest amount of protein.

A way to figure out the legitimacy of these assumptions is to calculate the average fat, carbohydrate, fiber and protein content of each item type.

```{r}
avg_by_type <- sbux %>%
  group_by(type) %>%
  summarize(avg_fat = mean(fat),
            avg_carb = mean(carb),
            avg_fiber = mean(fiber),
            avg_protein = mean(protein))
avg_by_type
```
We can plot each macronutrient in a separate bar graph:

```{r}
ggplot(avg_by_type, aes(x=reorder(type, -avg_fat), y=avg_fat, fill=type)) +
  geom_bar(stat='identity') +
  labs(title='Average Fat by Type', x='Type', y='Average Fat (g)')
```

```{r}
ggplot(avg_by_type, aes(x=reorder(type, -avg_carb), y=avg_carb, fill=type)) +
  geom_bar(stat='identity') +
  labs(title='Average Carb by Type', x='Type', y='Average Carb (g)')
```

```{r}
ggplot(avg_by_type, aes(x=reorder(type, -avg_fiber), y=avg_fiber, fill=type)) +
  geom_bar(stat='identity') +
  labs(title='Average Fiber by Type', x='Type', y='Average Fiber (g)')
```

```{r}
ggplot(avg_by_type, aes(x=reorder(type, -avg_protein), y=avg_protein, fill=type)) +
  geom_bar(stat='identity') +
  labs(title='Average Protein by Type', x='Type', y='Average Protein (g)')
```
Bakery items do indeed have more carbohydrates than other item types, but they are closely followed by the parfait type. Sandwiches also have the highest protein content than other types. When it comes to fat and fiber, the bistro box category has on average the highest amount of both.

# Are there any food items that are particularly rich in fiber but low in fat and carbohydrates?
To answer this question, we first need to define what counts as “rich in fiber” or “low in fat/carbs”. To do so, we can look at the summary statistics for each macronutrient.
```{r}
summary(sbux)
```
To determine what counts as “rich in fiber” we will take the 3rd quartile value of 4 grams. For fat, we are going to take the Q1 value of 9 grams. We will do the same for carbohydrates. Even though the Q1 carb value of 31 is not technically low, if we pick a lower value we might not be able to find any items that fit our criteria.
```{r}
q3 <- sbux %>%
  filter(fiber > 4, fat < 9, carb < 31) %>%
  select(item, fiber, fat, carb)
q3
```

# Correlation between Caloric Content and Macronutrient Amount

To find out if there is a correlation between the caloric content of Stabucks’ food items and their nutritional values, we can create a correlation matrix and then visualize its values using a heatmap.
```{r}
cor_matrix <- cor(sbux[, c("calories", "fat", "carb", "fiber", "protein")])
cor_matrix

# remove dendogram
heatmap(cor_matrix, Rowv = NA, Colv = NA)
```
Calories are highly correlated with fat and carb values in an item, but not so much with protein and even less with fiber.

# Trends and Patterns in Nutritional Composition, Other Outliers
So far in our analysis we have spotted a couple of trends, such as:

1. More than half of the food items at Starbucks belong to the bakery type.

2. The middle 50% of the food items have a caloric content between 300 and 420 kcal.

3. The higher the number of calories in an item, the more likely its fat and carbohydrate content is high as well.

4. Bakery items have the most carbs, sandwiches the most protein, and parfait items the most fat & fiber.

Something else we can look into is the distribution of macronutrients among all the food items.
```{r}
boxplot(sbux$fat, sbux$carb, sbux$fiber, sbux$protein,
        names = c("Fat", "Carbohydrates", "Fiber", "Protein"),
        main = "Nutritional Properties Box Plot")
```
From this box plot we can confirm that Starbucks food items are richer in carbohydrates than any other nutrient. However, this chart shows us that there are two outliers in the protein nutrient.
```{r}
p_out <- boxplot.stats(sbux$protein)$out
p_out_indices <- which(sbux$protein %in% p_out)
p_out_items <- starbucks[p_out_indices, "item"]
p_out_items
```

# Differences in the Nutritional Profiles of Food Items Across Different Categories

So far, we have observed that different food types have different average compositions. It is helpful to map these compositions in a stacked bar chart, to observe the differences across the types simultaneously.

```{r}
avg_composition <- aggregate(cbind(fat, carb, fiber, protein) ~ type, sbux, mean)

# Reshape the data into long format
avg_composition_long <- tidyr::gather(avg_composition, nutrient, value, -type)

# Create the stacked bar chart
compositions <- ggplot(avg_composition_long, aes(x = type, y = value, fill = nutrient)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Composition by Food Type", x = "Food Type", y = "Average Composition") +
  scale_fill_manual(values = c("steelblue", "lightgreen", "orange", "pink"), 
                    labels = c("Fat", "Carbohydrates", "Fiber", "Protein")) +
  theme_minimal()
compositions
```

# Food Items that Provide a Good Balance of Macronutrients

According to medical professionals (https://www.prospectmedical.com/resources/wellness-center/macronutrients-fats-carbs-protein#:~:text=In%20general%2C%20most%20adults%20should,30%2D40%25%20Fat.), most adults should target their diets to comprise of 45-65% carbohydrates, 10-35% protein and 20-35% fat. Keeping that in mind, we can determine which items from the Starbucks dataset can serve as a better meal option.

First, we need to figure out the composition of each item from the list:
```{r}
sum_composition <- aggregate(cbind(fat, carb, fiber, protein) ~ item, sbux, sum)

# Calculate the ratios for each macronutrient within each food type
ratio_composition <- transform(sum_composition, 
                               fat_ratio = round(fat / (fat + carb + fiber + protein), 3),
                               carb_ratio = round(carb / (fat + carb + fiber + protein), 3),
                               fiber_ratio = round(fiber / (fat + carb + fiber + protein), 3),
                               protein_ratio = round(protein / (fat + carb + fiber + protein), 3))

# Display the ratio composition
composition_ratios <- subset(ratio_composition, select = -c(fat, carb, fiber, protein))
head(composition_ratios)
```
Now that we have the list, we can pick out the items that have macronutrients spread across those percentages:

```{r}
balanced_food <- composition_ratios[
  composition_ratios$fat_ratio >= 0.2 & composition_ratios$fat_ratio <= 0.35 &
    composition_ratios$carb_ratio >= 0.45 &
    composition_ratios$carb_ratio <= 0.65 &
    composition_ratios$protein_ratio >= 0.1 & composition_ratios$protein_ratio <= 0.35,
]
balanced_food
```
Using the criteria specified at the beginning of this section, there are 6 items in the Starbucks food list that can be considered balanced meals in terms of macronutrient composition. All of them are sandwiches, with the exception of “Protein” that falls under bistro box.